---
title: "Descriptives"
author: "Caroline Ledbetter"
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=3.5, fig.height=3, 
                      echo=FALSE, warning=FALSE, message=FALSE)
```
```{r setup, include=FALSE}
load(file = "~/Repositories/Data/Capstone/descriptives.Rdata")
library(pander)
```

There are `r nrow(analysis)` subjects with a non-zero fasting glucose weight (included in the morning session for fasting glucose testing). `r length(which(analysis$exclreason == 'Age'))` observations were excluced for being outide the age range and `r length(which(analysis$exclreason == 'Pregnancy'))` observations were excluded for pregnancy for a total of `r length(which(analysis$subset == F))` exclusions. In addition there are `r length(which(is.na(analysis$syndrome)))` observations with missing information on metabolic syndrome. Number of observations missing each criteria are listed in table 1. Demographic and covariate information by food insecurity status are shown in Table 2 including information on those whose food insecurity status is missing. Table 3 shows metabolic syndrome by BMI category. Table 4 shows the unweighted table of food insecurity by metabolic syndrome status and Table 5 shows the number of individuals in each food security category. 

```{r table one}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(missingtable, caption = 'Missing Metabolic Syndrome Variables')
```


```{r table two}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(table1, caption = 'Characteristics of Study Participants by Food Insecurity Status')
```

```{r table three}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(Table3, caption = 'BMI Category by Metabolic Syndrome Status')
```

```{r table four}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(Table4, caption = 'Metabolic Syndrome Status by Food Insecurity')
```

```{r table five}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(t(table5), caption = 'No of Individuals at Each Level of Food Insecurity')
```

