---
title: "Association of Food Insecurity and Metabolic Syndrome among NHANES Participants 1999-2014"
author: "Caroline Ledbetter"

output: 
  pdf_document:
    fig_caption: yes
    fig_height: 3.5
    fig_width: 8

bibliography: /Users/Caroline/Repositories/Capstone-Project/Docs/paper.bib
csl: /Users/Caroline/Repositories/Capstone-Project/Docs/journal-of-nutrition.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r setup, include=FALSE}
load(file = "~/Repositories/Data/Capstone/finaldescriptives.rdata")
load(file = '~/Repositories/Data/Capstone/analysis_report.rdata')
library(pander)
library(ggplot2)
library(grid)
library(gridExtra)
```

# Abstract  

# Introduction  

Food Security is access at all times to enough food for an active, healthy life; food insecurity exists when ever the availability of nutritionally adequate and safe foods is limited or uncertain[@RN81]. In 2010, it was estimated that approximately 50 million Americans were food insecure [@RN82]. Houselholds reporting food insecurity report a decrease in the frequency of comnsumption of fruits and vegetables[@RN86]. Previous has research has found an association between food insecurity and hyperlipidemia, hypertension, diabetes[@RN80; @RN89], peripheral arterial disease [@RN88], and poor cardiovascular health [@RN87]. 

# Methods  

# Results   

\newpage
```{r table one}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(table1, caption = 'Charactersitcs of Study Participants by Food Security Status')
set.alignment('center', row.names = 'left')
pander(table1_alt, 
       caption = 'Charactersitcs of Study Participants by Food Security Category')
```

```{r table two}
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(table2, caption = 'Weighted Charactersitcs of Study Participants by Food Security Status')
set.alignment('center', row.names = 'left')
pander(table2alt, 
       caption = 'Weighted Charactersitcs of Study Participants by Food Security Category')
```

\newpage
```{r age-period graphs}
title2 <- 
  textGrob("Prevalence of Metabolic Syndrome and Food Insecurity by Age and NHANES Cycle", 
           gp=gpar(fontface="bold"))
grid.arrange(arrangeGrob(plotMSbyAGe + theme(legend.position="none"),
                         plotFIbyAGe + theme(legend.position="none"),
                         nrow=1, ncol = 2, bottom = 'Midpoint of Age'), 
             mylegend, ncol = 2, widths = c(6, 1), top = title2, left = 'Prevalence')

```

```{r foodseccat graphs, fig.height = 7}
title1=textGrob("Prevalence of Food Security Categories by Age and NHANES Cycle",
                gp=gpar(fontface="bold"))
grid.arrange(arrangeGrob(p_FS2 + theme(legend.position="none"),
                         p_FS3 + theme(legend.position="none"),
                         p_FS4 + theme(legend.position="none"),
                         nrow=3, ncol = 1),
              mylegend, ncol = 2, widths = c(6, 1), top = title1, left = 'Prevalence')
```
\newpage



```{r results}
# function to round ratios and put CIs in parentheses
CItxt <- function(obj) {
  y <- sapply(obj, function(x) sprintf('%.2f',x))
  z <- paste0(y[1], '(', y[2], '-', y[3], ')')
  return(z)
}

RR_basic <- CItxt(RR_basic)
RR_basic_noalc <- CItxt(RR_basic_noalc)

# unadjusted prevalence and Crude Ratios - Dichotimous Food Insecurity
UnadjTbl1 <- matrix(c('',prevalence, RR_basic, '',prevalence_noalc, RR_basic_noalc),
                     ncol = 2, 
                     dimnames = list(c('Unadjusted Prevalence',
                                       '\\ Food Secure',
                                       '\\ Food Insecure',
                                       'Crude Risk Ratio'), 
                                     c('Metabolic Syndrome (missing alc excl)', 
                                       'Metabolic Syndrome (missing alc not excl)')))
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(UnadjTbl1, 
       caption = 'Unadjusted Prevalence and Relative Risk Ratio of Metabolic Syndrome')


# unadjusted prevalence and Crude Ratios - Categorical Food Security
RR_basic2 <- apply(RR_basic2, MARGIN = 1, CItxt)
RR_basic_noalc2 <- apply(RR_basic_noalc2, MARGIN = 1, CItxt)

UnadjTbl2 <- matrix(c('', prevalence_alt, '', RR_basic2, '',prevalence_alt_noalc, 
                      '', RR_basic_noalc2), ncol = 2, 
                     dimnames = list(
                       c('Unadjusted Prevalence', '\\ Full food security', 
                         '\\ Marginal food security', 
                         '\\ Low Food security',
                         '\\ Very low Food security', 
                         'Crude Risk Ratio (vs full food security)',
                         '\\ Marginal food security',
                         '\\ Low food security',
                         '\\ Very low food security'), 
                       c('Metabolic Syndrome (missing alc excl)', 
                         'Metabolic Syndrome (missing alc not excl)')))
panderOptions('table.split.table', Inf)
set.alignment('center', row.names = 'left')
pander(UnadjTbl2, 
       caption = 'Unadjusted Prevalence and Relative Risk Ratio of Metabolic Syndrome')


# dicotimous food security with and without alcolhol
RR_male <- CItxt(RR_male)
RR_female <- CItxt(RR_female)
RR_male_noalc <- CItxt(RR_male_noalc)
RR_female_noalc <- CItxt(RR_female_noalc)

BinTab <- matrix(c(RR_male, RR_female, RR_male_noalc, RR_female_noalc), nrow = 2,
                 dimnames = list(c('Male', 'Female'), 
                                 c('Adj RR', 'Adj RR not adj for alc')))
set.alignment('center', row.names = 'left')
pander(BinTab, 
       caption = 'Adjusted Risk Ratio Metabolic Syndrome by Food Insecurity Status')


# Categorical Food Security Status
RR_male_cat <- apply(RR_male_cat, 1, CItxt)
RR_male_cat_noalc <- apply(RR_male_cat_noalc, 1, CItxt)

CatTab <- matrix(c('', RR_male_cat, '', RR_male_cat_noalc,
                   '', RR_female_cat, '', RR_female_cat_noalc), ncol = 2,
                 dimnames = list(c('Adj RR (vs full food security)', 
                                   '\\ Marginal food security', 
                                   '\\ Low food security', 
                                   '\\ Very low food security',
                                   'Adj RR not adj for alc (vs full food security)', 
                                   '\\ Marginal food security', 
                                   '\\ Low food security', 
                                   '\\ Very low food security'), 
                                 c('Males', 'Females')))
set.alignment('center', row.names = 'left')
pander(CatTab, 
       caption = 'Adjusted Risk Ratio Metabolic Syndrome by Food Security Category')
```



# Discussion  

# Conculsions  

# References


